---
title: "Project 3"
author: "Sciper 359523"
date: "2023-03-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
```

```{r working directory and data, echo=FALSE, include=FALSE}
lapply(c("dplyr","chron","ggplot2","tidyr","questionr","survival","forcats","tidyselect",
         "data.table","table1","lubridate", "ggpubr","viridis","finalfit","survminer",
         "ggpubr", "ggthemes", "gridExtra", "rstatix","stringr",
         "wesanderson","kableExtra", "naniar","boot","scales","ggsci", "stringr",
         "Hmisc","DescTools","swimplot", 'stats', 'EnvStats', 'finalfit'), 
       library, character.only=TRUE)

Sys.setlocale("LC_TIME", "English")

#data bases import
df18 <- read.csv(file = '2018-19.csv')
df19 <- read.csv(file = '2019-20.csv')
df20 <- read.csv(file = '2020-21.csv')
df21 <- read.csv(file = '2021-22.csv')


set.seed(435) #set the seed for reproducible results
```

# 1. Introduction
Home advantage is a well-known phenomenon in football where a team playing at their home stadium has an advantage over the away team. This advantage can be attributed to factors such as familiarity with the playing surface and surroundings, support of home fans, and a psychological boost. In addition, referees may also be influenced by the home crowd, which is known as refereeing bias. A study by Garicano and Palacios-Huerta (2001)[1] found that home teams in the Spanish La Liga received significantly fewer yellow and red cards than away teams, suggesting that referees may be influenced by the home crowd.

The magnitude of the home advantage effect has been highlighted by previous studies. Nevill, Newell, and Gale (1996) [2] analyzed over 10,000 matches from English football leagues and found that home teams won approximately 60% of matches, while away teams won only 20% of matches. A more recent study by Pollard [3] analyzed data from the UEFA Champions League and concluded that home teams won 62.4% of matches, while away teams won only 20.3% of matches.

In this project, we analyze the English Premier League during the seasons 2018-2019 to 2021-2022. Due to Covid-19, teams were forced to play in empty stadiums between March 2020 and the beginning of the 2021-22 season. This presents an opportunity to investigate the strength of the home advantage effect during closed-door matches, where neither the players nor the referees are influenced by the crowd. Thus, this report will present the most significant results concerning the effect of Covid-19 on home advantage.

In Section 2, we introduce the dataset, and in Section 3, we describe our model building procedure and present the final model. Section 4 analyzes our model via diagnostics, and we will analyze its stability in Section 5. In Section 6, we interpret and analyze the results, and finally, we conclude the work done and discuss potential improvements in Section 7.

# 2. Data characteristics
The data used for analysis consists of match results and statistics, betting odds, and bookmaker information from the English Premier League. The data set is a concatenation of four separate data sets, each containing match results for four seasons : 2018-2019, 2019-2020, 2020-2021, and 2021-2022. The match results variables include the winning team, home and away teams, goals scored by each team at half-time and full-time, and the number of shots and shots on target, as well as yellow and red cards for each team. In addition, the data set includes information on the number of BetBrain bookmakers used to calculate match odds averages and maximums, and total goals betting odds. As a first definition of home advantage, we primarily examined the impact of playing at home on a team's goal scoring ability. Secondly, we used the match results (win or loose) to build a poisson model.

```{r tidy the data and rename, echo=FALSE}
#merge the data sets together
df <- bind_rows(df18, df19, df20, df21)

#date
df <- df%>%mutate(
  Date = as.Date(Date , format = "%d/%m/%Y")
)

#restrict the number of variables
df_restr <- df %>% select(Date, HomeTeam, AwayTeam, FTR, FTHG, FTAG)

#pivot longer to create two rows for each match
df_restr <- df_restr %>% pivot_longer(cols = c(HomeTeam, AwayTeam), names_to = 'Team')

#variable win
df_restr <- df_restr%>% mutate(
  win = as.factor(case_when(
    Team == 'HomeTeam' & FTR == 'H' ~ 1, 
    Team == 'AwayTeam' & FTR == 'H' ~ 0,
    Team == 'HomeTeam' & FTR == 'A' ~ 0,
    Team == 'AwayTeam' & FTR == 'A' ~ 1,
    Team == 'HomeTeam' & FTR == 'D' ~ 2,
    Team == 'AwayTeam' & FTR == 'D' ~ 2,
    
  )
))

#score variable 
df_restr <- df_restr %>% mutate(
  goals = case_when(
    Team == 'HomeTeam' ~ FTHG,
    Team == 'AwayTeam' ~ FTAG
  )
)

#variable to assess if covid or not
df_restr <- df_restr %>% mutate(
  covid = as.factor(case_when(
      Date >= "2020-03-12" & Date <= "2021-06-01"~ 1,
      TRUE ~ 0
  ))
)

df_restr <- df_restr %>% mutate(
  covid_before_after = as.factor(case_when(
      Date >= "2020-03-12" & Date <= "2021-06-01"~ "In-covid",
      Date <= "2020-03-12" ~"Pre-covid",
      TRUE ~ "Post-covid"
  ))
)

#rename variables
df_restr <- df_restr %>% mutate(
  home_goals = FTHG,
  away_goals = FTAG,
  team_name = value,
  team_status = Team,
  covid_y_n = covid,
  date = Date
)

#restricted data set with only the variables of interest
df_restr <- df_restr %>% 
  select(date, home_goals, away_goals, team_name, team_status, covid_y_n, covid_before_after, win, goals )

#data frame with only the values before covid
df_before <- df_restr %>%
  filter(covid_before_after == 'Pre-covid')

#data frame with only the values during covid
df_during <- df_restr %>% 
  filter(covid_before_after == 'In-covid')

#data frame with only the values after covid
df_after <- df_restr %>% 
  filter(covid_before_after == 'Post-covid')

#data frames that count the number of victories before covid and all periods of time 
df_wins_before <- df_restr %>% 
  filter(covid_before_after =='Pre-covid')%>%  
  filter(win == "1")%>% 
  select(team_name, team_status, win)%>% 
  group_by(team_name, team_status)%>%
  summarise(number=n())%>% 
  mutate(numb_win = number)

df_wins_all <- df_restr %>% 
  select(team_name, team_status, win)%>% 
  filter(win == "1") %>% 
  group_by(team_name, team_status)%>%
  summarise(number=n())%>% 
  mutate(numb_win = number)

df_wins_during <- df_restr %>%
  filter(covid_before_after =='In-covid')%>%  
  filter(win == "1")%>% 
  select(team_name, team_status, win)%>% 
  group_by(team_name, team_status)%>%
  summarise(number=n())%>% 
  mutate(numb_win = number)

df_wins_after <- df_restr %>% 
  filter(covid_before_after =='Post-covid')%>%  
  filter(win == "1")%>% 
  select(team_name, team_status, win)%>% 
  group_by(team_name, team_status)%>%
  summarise(number=n())%>% 
  mutate(numb_win = number)


```

*Figure1* illustrates the difference in goals between the home team and the away team over time. The red line represents the average of the series, while the blue zone highlights the period during the COVID-19 pandemic when matches were played without spectators in attendance. This period is commonly referred to as closed-doors matches. The graph provides insights into how the home advantage phenomenon might be affected by playing in empty stadiums compared to games where fans are present.

```{r, fig.align="center", fig.cap="Figure 1 : Difference between home goals and away goals"}
graph<- df_restr %>% 
  select(date, home_goals, away_goals) %>% 
  group_by(date) %>% 
  summarise(avg_home = mean(home_goals), avg_away = mean(away_goals)) %>%
  mutate(avg_diff = avg_home - avg_away) 

graph %>% ggplot(aes(x=date, y=avg_diff)) + 
  geom_line() + 
  geom_hline(yintercept= mean(graph$avg_diff), col = 'red', show.legend = T)+ 
  geom_rect(aes(xmin = as.Date("2020-03-12", format = "%Y-%m-%d"), xmax = as.Date("2021-06-01", format = "%Y-%m-%d"), ymin = -Inf, ymax = Inf), fill = "blue", alpha = 0.003) +
  labs(x = 'Date', y = 'Average difference of goals', title = 'Difference between home goals and away goals')


```

The figure indicates that there is a slight advantage to playing at home since the average difference in the number of goals scored is positive. It appears that the COVID-19 pandemic had little impact on the average difference in the number of goals scored, although there are fewer extreme scores during the closed-doors matches. In the next section, we will model the data to gain a deeper understanding of the extent to which the home advantage persisted during the COVID-19 pandemic. Our analysis aims to provide more robust evidence on the impact of playing without fans and whether it affected the home advantage in the English Premier League.

# 3. Model building

## a. Linear model
Home advantage can manifest in various forms, and we consider one indicator to be a team scoring more goals when playing at home. We constructed separate linear models for each COVID-19 period (before, during, and after) to evaluate the effect of playing at home on a team's goal scoring ability. Specifically, we assessed whether there was a statistically significant difference in the number of goals scored by a team when playing at home compared to away during each period. 

```{r linear model, echo=FALSE, include = FALSE}
#model before covid
linear_model_before <- lm(goals ~ team_status, data = df_before)
summary(linear_model_before)

#during covid 
linear_model_during <- lm(goals ~ team_status , data = df_during)
summary(linear_model_during)

#after covid 
linear_model_after <- lm(goals ~ team_status , data = df_after)
summary(linear_model_after)

#compare before and after covid
linear_model_b_a <- lm(goals ~ team_status + covid_before_after, data = df_restr)
summary(linear_model_b_a)

```

The estimated coefficient associated with the home team is given in the following table :

*Table1 : Estimated coefficients with the linear model*


|              | Estimate | Std.Error | t value | Pr(>\|t\|) |2.5% | 97.5%|
|--------------|----------|-----------|---------|------------|-----|------|
| Before covid | 0.27     | 0.069     | 3.93    | 9.04e-05   |0.14 | 0.4  | 
| During covid | 0.01     | 0.09      | 0.11    | 0.91       |-0.17| 0.19 |
| After covid  | 0.20     | 0.09      | 2.22    | 0.03       |0.02 | 0.39 |

The equation for this model can be written 
$N_{goals} = \beta*Home + \epsilon$

where $\beta$ is the estimate, $\epsilon$ is the residual, $Home$ denotes the fact that the teams plays home and $N_{goals}$ is the number of goals.

## b. Poisson model
Another way to model the home advantage is through the use of a Poisson model, also known as a log-linear model. Here, we explore whether home teams have an advantage in terms of winning matches. To this end, we consider the same three COVID-19 periods (before, during, and after) and estimate the coefficient associated with playing at home on our count data. This approach provides an alternative perspective on the home advantage phenomenon, which can help us to gain a more comprehensive understanding of the impact of playing at home in the English Premier League.

```{r poisson model, echo=FALSE, include = FALSE}
model_wins_before <- glm(df_wins_before$number ~ df_wins_before$team_status, family=poisson(link="log"))
summary(model_wins_before)

model_wins_during <- glm(df_wins_during$number ~ df_wins_during$team_status, family=poisson(link="log"))
summary(model_wins_during)

model_wins_after <- glm(df_wins_after$number ~ df_wins_after$team_status, family=poisson(link="log"))
summary(model_wins_after)
```

Table 2 displays the results.

*Table2 : Estimated coefficients with the Poisson regression*


|              | Estimate | Std.Error | t value | Pr(>\|t\|) | 2.5%   | 97.5% |
|--------------|----------|-----------|---------|------------|--------|-------|
| Before covid | 0.32     | 0.09      | 3.57    | 0.0003     | 0.15   | 0.5   |
| During covid | -0.06    | 0.12      | -0.52   | 0.6        | -0.28  | 0.17  |
| After covid  | 0.23     | 0.12      | 1.99    | 0.047      | 0.004  | 0.47  |

The equation for this model can be written 
$log(\mathbb{E}(N_{goals})) = \beta*Home + \epsilon$

where $\beta$ is the estimate, $\epsilon$ is the residual, $Home$ denotes the fact that the teams plays home and $N_{goals}$ is the number of goals.


# 4. Residuals diagnostics
## a. During 2019-20 season
We fit the previously mentioned model only for the 2019-2020 season, which was the most affected by the COVID-19 pandemic. The model diagnostics are presented in *Figure 2*. The residual plot displays two distinct bands, which correspond to the two values of the variable "team status" : home team (playing at home) and away team (playing away). The normal QQ-plot indicates that the residuals are skewed with some outliers. It may be worthwhile to refit the model without these influential outliers to determine the extent to which the model is impacted. The scale-location and constant leverage plots also identify these outliers.


```{r, fig.align="center", fig.cap="Figure 2 : Residuals diagnostics for the model with only the 2019-20 season"}
#restrict the number of variables
df_restr_19 <- df %>% select(Date, HomeTeam, AwayTeam, FTR, FTHG, FTAG)

#pivot longer to create two rows for each match
df_restr_19 <- df_restr_19 %>% pivot_longer(cols = c(HomeTeam, AwayTeam), names_to = 'Team')

#variable win
df_restr_19 <- df_restr_19 %>% mutate(
  win = as.factor(case_when(
    Team == 'HomeTeam' & FTR == 'H' ~ 1, 
    Team == 'AwayTeam' & FTR == 'H' ~ 0,
    Team == 'HomeTeam' & FTR == 'A' ~ 0,
    Team == 'AwayTeam' & FTR == 'A' ~ 1,
    Team == 'HomeTeam' & FTR == 'D' ~ 2,
    Team == 'AwayTeam' & FTR == 'D' ~ 2,
    
  )
))

#score variable 
df_restr_19 <- df_restr_19 %>% mutate(
  goals = case_when(
    Team == 'HomeTeam' ~ FTHG,
    Team == 'AwayTeam' ~ FTAG
  )
)

#variable to assess if covid or not
df_restr_19 <- df_restr_19 %>% mutate(
  covid = as.factor(case_when(
      Date >= "2020-03-12" & Date <= "2021-06-01"~ 1,
      TRUE ~ 0
  ))
)

df_restr_19 <- df_restr_19 %>% mutate(
  covid_before_after = as.factor(case_when(
      Date >= "2020-03-12" & Date <= "2021-06-01"~ "In-covid",
      Date <= "2020-03-12" ~"Pre-covid",
      TRUE ~ "Post-covid"
  ))
)

#rename variables
df_restr_19 <- df_restr_19 %>% mutate(
  home_goals = FTHG,
  away_goals = FTAG,
  team_name = value,
  team_status = Team,
  covid_y_n = covid,
  date = Date
)

#final data set
df_restr_19 <- df_restr_19 %>% select(date, home_goals, away_goals, team_name, team_status, covid_y_n, covid_before_after, win, goals )

df_restr_19 <- df_restr_19%>%select(team_name, team_status, win) %>% filter(win == "1")%>% group_by(team_name, team_status)%>%summarise(number=n())%>% mutate(numb_win = number)

model_19 <- glm(df_restr_19$number ~ df_restr_19$team_status, family=poisson(link="log"))


#res <- residuals(model_19, type="deviance")
par(mfrow=c(2,2))
plot(model_19)
```



## b. All seasons together
By fitting the model for all the seasons together (which captures an overall positive effect of playing at home, even though the effect is not present during the COVID-19 pandemic), we obtain the diagnostics displayed in *Figure 3*. The diagnostic plots have a similar shape to those for the model fitted only for the 2019-2020 season. Therefore, we once again identify some influential outliers, mostly located on the top right of the QQ-plot. These outliers may have a significant impact on the model, and it may be necessary to investigate them further to determine if they should be removed or if they provide important information.

```{r, fig.align="center", fig.cap="Figure 3 : Residuals diagnostics for the model considering all seasons"}
par(mfrow=c(2,2))
model_wins <- glm(df_wins_all$number ~ df_wins_all$team_status, family=poisson(link="log"))
plot(model_wins)
```   

# 5. Stability 
## a. Promoted and relegated teams
The models we have built so far have not taken into account the fact that some teams are relegated and others are promoted from season to season, which could potentially impact the estimates of the coefficients. To address this, we discard those teams and refit the Poisson model described in Part 5.b. This new model takes into account only the teams that have participated in all four seasons, and the estimated coefficients are based on these teams only. The estimated coefficients for the remaining teams are presented in the following table:

```{r promoted, echo=FALSE, include = FALSE}
#relegated/promoted effect
A = setdiff(df18$HomeTeam,df19$HomeTeam) #relegated (those in for 18 but not in 19)
B = setdiff(df19$HomeTeam,df18$HomeTeam) #promoted (those in 19 but not in 18)
C = setdiff(df19$HomeTeam,df20$HomeTeam) #relegated
D = setdiff(df20$HomeTeam,df19$HomeTeam) #promoted
E = setdiff(df20$HomeTeam,df21$HomeTeam ) #relegated
f = setdiff(df21$HomeTeam,df20$HomeTeam ) #promoted

names_del <- unique(c(A, B, C, D, E, f)) #we will not take in account the ones that were relegated and promoted

df_wins_before_rel <- df_restr %>% filter(covid_before_after =='Pre-covid')%>%  filter(win == "1")%>% filter(!(team_name%in%names_del))%>%select(team_name, team_status, win)%>% group_by(team_name, team_status)%>%summarise(number=n())%>% mutate(numb_win = number)

df_wins_during_rel <- df_restr %>% filter(covid_before_after =='In-covid')%>%  filter(win == "1")%>% filter(!(team_name%in%names_del))%>%select(team_name, team_status, win)%>% group_by(team_name, team_status)%>%summarise(number=n())%>% mutate(numb_win = number)

df_wins_after_rel <- df_restr %>% filter(covid_before_after =='Post-covid')%>%  filter(win == "1")%>% filter(!(team_name%in%names_del))%>%select(team_name, team_status, win)%>% group_by(team_name, team_status)%>%summarise(number=n())%>% mutate(numb_win = number)

model_wins_before_rel <- glm(df_wins_before_rel$number ~ df_wins_before_rel$team_status, family=poisson(link="log"))
summary(model_wins_before_rel)

model_wins_during_rel <- glm(df_wins_during_rel$number ~ df_wins_during_rel$team_status, family=poisson(link="log"))
summary(model_wins_during_rel)

model_wins_after_rel <- glm(df_wins_after_rel$number ~ df_wins_after_rel$team_status, family=poisson(link="log"))
summary(model_wins_after_rel)

```

*Table 3 : Estimated coefficients of the Poisson model without relegated and promoted teams*

|              | Estimate | Std.Error | t value | Pr(>\|t\|) |
|--------------|----------|-----------|---------|------------|
| Before covid | 0.28     | 0.1       | 2.86    | 0.004      |
| During covid | -0.07    | 0.13      | -0.51   | 0.61       |
| After covid  | 0.29     | 0.13      | 2.29    | 0.02       |



## b. Draw effect
Another limitation of the model described in Part 5.b is that we only counted the number of wins, but we could also include the draws as a "victory" for both teams. By doing this, we can account for the so-called "draw effect", which is the tendency of home teams to draw more often than away teams. This effect could potentially impact the estimates of the coefficients. To address this issue, we refit the Poisson model by counting draws as 0.5 wins for both teams. The estimated coefficients for this modified model are presented in the table below :


```{r draw effect, echo=FALSE, include = FALSE}
#if we consider that draw is like winning 
df_wins_before_bis <- df_restr %>%
  filter(covid_before_after =='Pre-covid')%>% 
  filter(win != "0")%>% 
  select(team_name, team_status, win)%>% 
  group_by(team_name, team_status)%>%
  summarise(number=n())%>% 
  mutate(numb_win = number)

df_wins_during_bis <- df_restr %>% 
  filter(covid_before_after =='In-covid')%>%  
  filter(win != "0")%>%
  select(team_name, team_status, win)%>% 
  group_by(team_name, team_status)%>%
  summarise(number=n())%>%
  mutate(numb_win = number)

df_wins_after_bis <- df_restr %>% 
  filter(covid_before_after =='Post-covid')%>%  
  filter(win != "0")%>% 
  select(team_name, team_status, win)%>%
  group_by(team_name, team_status)%>%
  summarise(number=n())%>%
  mutate(numb_win = number)


model_wins_before_bis <- glm(df_wins_before_bis$number ~ df_wins_before_bis$team_status, family=poisson(link="log"))
summary(model_wins_before_bis)

model_wins_during_bis <- glm(df_wins_during_bis$number ~ df_wins_during_bis$team_status, family=poisson(link="log"))
summary(model_wins_during_bis)

model_wins_after_bis <- glm(df_wins_after_bis$number ~ df_wins_after_bis$team_status, family=poisson(link="log"))
summary(model_wins_after_bis)




#if we consider draw as half victory
df_wins_before_draw <- df_restr %>% 
  filter(covid_before_after =='Pre-covid')%>%
  select(team_name, team_status, win)%>%
  mutate(
    win_numeric = case_when(
      win == 0 ~ 0,
      win == 1 ~ 1,
      win == 2 ~ 0.5
    )
  ) %>%
  group_by(team_name, team_status, win_numeric)%>%
  summarise(number = n(),s = sum(win_numeric)) %>% 
  group_by(team_name, team_status)%>%
  summarise(score = sum(s))

df_wins_during_draw <- df_restr %>% 
  filter(covid_before_after =='In-covid')%>%
  select(team_name, team_status, win)%>%
  mutate(
    win_numeric = case_when(
      win == 0 ~ 0,
      win == 1 ~ 1,
      win == 2 ~ 0.5
    )
  ) %>%
  group_by(team_name, team_status, win_numeric)%>%
  summarise(number = n(),s = sum(win_numeric)) %>% 
  group_by(team_name, team_status)%>%
  summarise(score = sum(s))

df_wins_after_draw <- df_restr %>% 
  filter(covid_before_after =='Post-covid')%>%
  select(team_name, team_status, win)%>%
  mutate(
    win_numeric = case_when(
      win == 0 ~ 0,
      win == 1 ~ 1,
      win == 2 ~ 0.5
    )
  ) %>%
  group_by(team_name, team_status, win_numeric)%>%
  summarise(number = n(),s = sum(win_numeric)) %>% 
  group_by(team_name, team_status)%>%
  summarise(score = sum(s))


model_wins_before_draw <- glm(df_wins_before_draw$score ~ df_wins_before_draw$team_status, family=poisson(link="log"))
summary(model_wins_before_draw)

model_wins_during_draw <- glm(df_wins_during_draw$score ~ df_wins_during_draw$team_status, family=poisson(link="log"))
summary(model_wins_during_draw)

model_wins_after_draw <- glm(df_wins_after_draw$score ~ df_wins_after_draw$team_status, family=poisson(link="log"))
summary(model_wins_after_draw)

#remark : if we take in account that draw is equal to victory we obtain quite similar results but the pvalues are a bit higher
```


*Table 4 : Estimated coefficients of the Poisson model taking in account the draws*

|              | Estimate | Std.Error | t value | Pr(>\|t\|) |
|--------------|----------|-----------|---------|------------|
| Before covid | 0.25     | 0.08      | 3.19    | 0.001      |
| During covid | -0.05    | 0.10      | -0.46   | 0.64       |
| After covid  | 0.18     | 0.10      | 1.74    | 0.08       |



# 6. Interpretation
In section 3.a, we present two models to quantify the home advantage. The first model examines the difference in the number of goals between playing at home and playing away. Before the COVID-19 pandemic, we observe a significant positive coefficient associated with playing at home, indicating a clear home advantage. The coefficient is practically significant, with a value of 0.27 (95CI = [0.14, 0.4]), which means that teams score 0.27 more goals when playing at home. However, during the pandemic, the coefficient associated with playing at home is not statistically significant since 0 belongs to the confidence interval, indicating no significant effect of playing at home. After the closed-door period, the effect of playing at home becomes statistically and practically significant again.

The second approach considers the number of strictly won games and models it using a Poisson or log-linear regression. The estimated coefficient before the pandemic is also statistically and practically significant (0.32) at the usual levels (5% and 1%). Again, the coefficient associated with the pandemic period is not significant, indicating no home advantage during this time. Finally, the home advantage reappears after the opening of stadiums, as observed in the linear regression model.



# 7. Discussion
The models developed provide a plausible explanation for the decrease in home advantage during the COVID-19 pandemic due to the closure of public stadiums. The diagnostic and stability analyses demonstrate the robustness of the models, which instills confidence in the results.

The most interesting finding of our work is that after the pandemic, the home advantage returned and did not disappear. The magnitude of the effect was generally similar but slightly smaller. Possible explanations for this decrease include teams becoming accustomed to playing in empty stadiums or changes in fan behavior before and after the pandemic.

Although we did not explore the effects of yellow and red cards on home advantage, which was studied by Garicano and Palacios-Huertas [1], this could be an interesting avenue for future research. Additionally, while we analyzed the number of goals in the linear regression model, it would be interesting to investigate the number of targeted goals instead.

# 8. References
[1] Garicano, L. and Palacios-Huerta, I. (2001). "An empirical examination of multilateral bargaining." American Economic Review, 91(4): 828-848.

[2] Nevill, A. M., Newell, S. M., & Gale, S. (1996). Factors associated with home advantage in English and Scottish soccer matches. Journal of Sports Sciences, 14(2), 181-186.

[3] Pollard, R., & Pollard, G. (2005). Home advantage in soccer: Variations in its magnitude and a literature review of the inter-related factors associated with its existence. Journal of Sport Behavior, 28(3), 169-189.