---
title: "Project 3"
author: "Sciper 359523"
date: "2023-03-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
```

```{r working directory and data, echo=FALSE, include=FALSE}
lapply(c("dplyr","chron","ggplot2","tidyr","questionr","survival","forcats","tidyselect",
         "data.table","table1","lubridate", "ggpubr","viridis","finalfit","survminer",
         "ggpubr", "ggthemes", "gridExtra", "rstatix","stringr",
         "wesanderson","kableExtra", "naniar","boot","scales","ggsci", "stringr",
         "Hmisc","DescTools","swimplot", 'stats', 'EnvStats', 'finalfit'), 
       library, character.only=TRUE)

Sys.setlocale("LC_TIME", "English")

#data bases import
df18 <- read.csv(file = '2018-19.csv')
df19 <- read.csv(file = '2019-20.csv')
df20 <- read.csv(file = '2020-21.csv')
df21 <- read.csv(file = '2021-22.csv')


set.seed(435) #set the seed for reproducible results
```

# 1. Introduction
Home advantage is a well-known phenomenon in football where a team playing at their home stadium has an advantage over the away team. This advantage can be attributed to factors such as familiarity with the playing surface and surroundings, support of home fans, and a psychological boost. In addition, referees may also be influenced by the home crowd, which is known as refereeing bias. A study by Garicano and Palacios-Huerta (2001)[1] found that home teams in the Spanish La Liga received significantly fewer yellow and red cards than away teams, suggesting that referees may be influenced by the home crowd.

The magnitude of the home advantage effect has been highlighted by previous studies. Nevill, Newell, and Gale (1996) [2] analyzed over 10,000 matches from English football leagues and found that home teams won approximately 60% of matches, while away teams won only 20% of matches. A more recent study by Pollard [3] analyzed data from the UEFA Champions League and concluded that home teams won 62.4% of matches, while away teams won only 20.3% of matches.

In this project, we analyze the English Premier League during the seasons 2018-2019 to 2021-2022. Due to Covid-19, teams were forced to play in empty stadiums between March 2020 and the beginning of the 2021-22 season. This presents an opportunity to investigate the strength of the home advantage effect during closed-door matches, where neither the players nor the referees are influenced by the crowd. Thus, this report will present the most significant results concerning the effect of Covid-19 on home advantage.

In Section 2, we introduce the dataset, and in Section 3, we describe our model building procedure and present the final model. Section 4 analyzes our model via diagnostics, and we will analyze its stability in Section 5. In Section 6, we interpret and analyze the results, and finally, we conclude the work done and discuss potential improvements in Section 7.

# 2. Data characteristics
The data on which the analyzes are made is composed of four data sets that were concatenated together. They all contain the match results for the four English premier league seasons :

+2018-2019
+2019-2020
+2020-2021
+2021-2022

The available variables contain match results (winning team, home and away team, goals from each team at half time and full time), match statistics (number of shots, number of shots on target and yellow and red cards for each team), betting odds, number of BetBrain bookmakers used to calculate match odds averages and maximums, and total goals betting odds. 

The variable that we will mostly use are the ones regarding match results. Indeed, we will make model regarding different definitions of the home advantage :

+the fact that a team is more or less likely to win if they play at home
+the fact that a team will goal more or less if they play at home 

```{r tidy the data and rename, echo=FALSE}
#merge the data sets together
df <- bind_rows(df18, df19, df20, df21)

#date
df <- df%>%mutate(
  Date = as.Date(Date , format = "%d/%m/%Y")
)

#restrict the number of variables
df_restr <- df %>% select(Date, HomeTeam, AwayTeam, FTR, FTHG, FTAG)

#pivot longer to create two rows for each match
df_restr <- df_restr %>% pivot_longer(cols = c(HomeTeam, AwayTeam), names_to = 'Team')

#variable win
df_restr <- df_restr%>% mutate(
  win = as.factor(case_when(
    Team == 'HomeTeam' & FTR == 'H' ~ 1, 
    Team == 'AwayTeam' & FTR == 'H' ~ 0,
    Team == 'HomeTeam' & FTR == 'A' ~ 0,
    Team == 'AwayTeam' & FTR == 'A' ~ 1,
    Team == 'HomeTeam' & FTR == 'D' ~ 2,
    Team == 'AwayTeam' & FTR == 'D' ~ 2,
    
  )
))

#score variable 
df_restr <- df_restr %>% mutate(
  goals = case_when(
    Team == 'HomeTeam' ~ FTHG,
    Team == 'AwayTeam' ~ FTAG
  )
)

#variable to assess if covid or not
df_restr <- df_restr %>% mutate(
  covid = as.factor(case_when(
      Date >= "2020-03-12" & Date <= "2021-06-01"~ 1,
      TRUE ~ 0
  ))
)

df_restr <- df_restr %>% mutate(
  covid_before_after = as.factor(case_when(
      Date >= "2020-03-12" & Date <= "2021-06-01"~ "In-covid",
      Date <= "2020-03-12" ~"Pre-covid",
      TRUE ~ "Post-covid"
  ))
)

#rename variables
df_restr <- df_restr %>% mutate(
  home_goals = FTHG,
  away_goals = FTAG,
  team_name = value,
  team_status = Team,
  covid_y_n = covid,
  date = Date
)

#restricted data set with only the variables of interest
df_restr <- df_restr %>% 
  select(date, home_goals, away_goals, team_name, team_status, covid_y_n, covid_before_after, win, goals )

#data frame with only the values before covid
df_before <- df_restr %>%
  filter(covid_before_after == 'Pre-covid')

#data frame with only the values during covid
df_during <- df_restr %>% 
  filter(covid_before_after == 'In-covid')

#data frame with only the values after covid
df_after <- df_restr %>% 
  filter(covid_before_after == 'Post-covid')

#data frames that count the number of victories before covid and all periods of time 
df_wins_before <- df_restr %>% 
  filter(covid_before_after =='Pre-covid')%>%  
  filter(win == "1")%>% 
  select(team_name, team_status, win)%>% 
  group_by(team_name, team_status)%>%
  summarise(number=n())%>% 
  mutate(numb_win = number)

df_wins_all <- df_restr %>% 
  select(team_name, team_status, win)%>% 
  filter(win == "1") %>% 
  group_by(team_name, team_status)%>%
  summarise(number=n())%>% 
  mutate(numb_win = number)

df_wins_during <- df_restr %>%
  filter(covid_before_after =='In-covid')%>%  
  filter(win == "1")%>% 
  select(team_name, team_status, win)%>% 
  group_by(team_name, team_status)%>%
  summarise(number=n())%>% 
  mutate(numb_win = number)

df_wins_after <- df_restr %>% 
  filter(covid_before_after =='Post-covid')%>%  
  filter(win == "1")%>% 
  select(team_name, team_status, win)%>% 
  group_by(team_name, team_status)%>%
  summarise(number=n())%>% 
  mutate(numb_win = number)


```

*Figure1* shows the difference of goals between the home goals and the away goals with respect to the date. The red line shows the average of the series and the blue zone represents the covid-19 period of time for which matches were played closed-doors.

```{r, fig.align="center", fig.cap="Figure 1 : Difference between home goals and away goals"}
test_2<- df_restr %>% 
  select(date, home_goals, away_goals) %>% 
  group_by(date) %>% 
  summarise(avg_home = mean(home_goals), avg_away = mean(away_goals)) %>%
  mutate(avg_diff = avg_home - avg_away) 

test_2 %>% ggplot(aes(x=date, y=avg_diff)) + 
  geom_line() + 
  geom_hline(yintercept= mean(test_2$avg_diff), col = 'red')+ 
  geom_rect(aes(xmin = as.Date("2020-03-12", format = "%Y-%m-%d"), xmax = as.Date("2021-06-01", format = "%Y-%m-%d"), ymin = -Inf, ymax = Inf), fill = "blue", alpha = 0.003) + 
  labs(x = 'Date', y = 'Average difference of goals', title = 'Difference between home goals and away goals')


```

The figure shows a slight advantage of playing home since the average is positive. It doesn't look like the covid-19 has a great impact on the average difference of number of goals, but there are less extremes scores. 

In the next section, we model the data to better understand what is happening and to what extent the home advantage persisted during covid-19. 

# 3. Model building

## a. Linear model
Home advantage can be seen through different indicators. As a first approach, we consider that a team has an advantage if they score more goals when they are at home. We build a linear model for each covid state (before, during, after) and evaluate in each state the effect of playing home.

```{r linear model, echo=FALSE, include = FALSE}
#model before covid
linear_model_before <- lm(goals ~ team_status, data = df_before)
summary(linear_model_before)

#during covid 
linear_model_during <- lm(goals ~ team_status , data = df_during)
summary(linear_model_during)

#after covid 
linear_model_after <- lm(goals ~ team_status , data = df_after)
summary(linear_model_after)

#compare before and after covid
linear_model_b_a <- lm(goals ~ team_status + covid_before_after, data = df_restr)
summary(linear_model_b_a)

```

The estimated coefficient associated with the home team is given in the following table :

*Table1 : Estimated coefficients with the linear model*


|              | Estimate | Std.Error | t value | Pr(>\|t\|) |2.5% | 97.5%|
|--------------|----------|-----------|---------|------------|-----|------|
| Before covid | 0.27     | 0.069     | 3.93    | 9.04e-05   |0.14 | 0.4  | 
| During covid | 0.01     | 0.09      | 0.11    | 0.91       |-0.17| 0.19 |
| After covid  | 0.20     | 0.09      | 2.22    | 0.03       |0.02 | 0.39 |

The equation for this model can be written 
$N_{goals} = \beta*Home + \epsilon$

where $\beta$ is the estimate, $\epsilon$ is the residual, $Home$ denotes the fact that the teams plays home and $N_{goals}$ is the number of goals.

## b. Poisson model
Another way of modelling the home advantage is to use a Poisson model (or log linear model). We now consider that a team has an advantage if they win more when they play at home. Again we consider the three phases, before, during and after covid-19 and estimate the coefficient associated with playing home.

```{r poisson model, echo=FALSE, include = FALSE}
model_wins_before <- glm(df_wins_before$number ~ df_wins_before$team_status, family=poisson(link="log"))
summary(model_wins_before)

model_wins_during <- glm(df_wins_during$number ~ df_wins_during$team_status, family=poisson(link="log"))
summary(model_wins_during)

model_wins_after <- glm(df_wins_after$number ~ df_wins_after$team_status, family=poisson(link="log"))
summary(model_wins_after)
```

Table 2 displays the results.

*Table2 : Estimated coefficients with the Poisson regression*


|              | Estimate | Std.Error | t value | Pr(>\|t\|) | 2.5%   | 97.5% |
|--------------|----------|-----------|---------|------------|--------|-------|
| Before covid | 0.32     | 0.09      | 3.57    | 0.0003     | 0.15   | 0.5   |
| During covid | -0.06    | 0.12      | -0.52   | 0.6        | -0.28  | 0.17  |
| After covid  | 0.23     | 0.12      | 1.99    | 0.047      | 0.004  | 0.47  |

The equation for this model can be written 
$log(\mathbb{E}(N_{goals})) = \beta*Home + \epsilon$

where $\beta$ is the estimate, $\epsilon$ is the residual, $Home$ denotes the fact that the teams plays home and $N_{goals}$ is the number of goals.


# 4. Residuals diagnostics
## a. During 2019-20 season
We can fit the previous model only for the 2019-2020 season which is the most impacted by covid-19. The diagnostics we obtain are presented in *Figure 2*. The residuals have a strange shape which could mean the model is not adequate. The normal QQ-plot shows that the residuals are skewed with some outliers. One could refit the model without those influential outliers to see to what extent the model is changed. The scale location and constant leverage also enable to identify those outliers.


```{r, fig.align="center", fig.cap="Figure 2 : Residuals diagnostics for the model with only the 2019-20 season"}
#restrict the number of variables
df_restr_19 <- df %>% select(Date, HomeTeam, AwayTeam, FTR, FTHG, FTAG)

#pivot longer to create two rows for each match
df_restr_19 <- df_restr_19 %>% pivot_longer(cols = c(HomeTeam, AwayTeam), names_to = 'Team')

#variable win
df_restr_19 <- df_restr_19 %>% mutate(
  win = as.factor(case_when(
    Team == 'HomeTeam' & FTR == 'H' ~ 1, 
    Team == 'AwayTeam' & FTR == 'H' ~ 0,
    Team == 'HomeTeam' & FTR == 'A' ~ 0,
    Team == 'AwayTeam' & FTR == 'A' ~ 1,
    Team == 'HomeTeam' & FTR == 'D' ~ 2,
    Team == 'AwayTeam' & FTR == 'D' ~ 2,
    
  )
))

#score variable 
df_restr_19 <- df_restr_19 %>% mutate(
  goals = case_when(
    Team == 'HomeTeam' ~ FTHG,
    Team == 'AwayTeam' ~ FTAG
  )
)

#variable to assess if covid or not
df_restr_19 <- df_restr_19 %>% mutate(
  covid = as.factor(case_when(
      Date >= "2020-03-12" & Date <= "2021-06-01"~ 1,
      TRUE ~ 0
  ))
)

df_restr_19 <- df_restr_19 %>% mutate(
  covid_before_after = as.factor(case_when(
      Date >= "2020-03-12" & Date <= "2021-06-01"~ "In-covid",
      Date <= "2020-03-12" ~"Pre-covid",
      TRUE ~ "Post-covid"
  ))
)

#rename variables
df_restr_19 <- df_restr_19 %>% mutate(
  home_goals = FTHG,
  away_goals = FTAG,
  team_name = value,
  team_status = Team,
  covid_y_n = covid,
  date = Date
)

#final data set
df_restr_19 <- df_restr_19 %>% select(date, home_goals, away_goals, team_name, team_status, covid_y_n, covid_before_after, win, goals )

df_restr_19 <- df_restr_19%>%select(team_name, team_status, win) %>% filter(win == "1")%>% group_by(team_name, team_status)%>%summarise(number=n())%>% mutate(numb_win = number)

model_19 <- glm(df_restr_19$number ~ df_restr_19$team_status, family=poisson(link="log"))


#res <- residuals(model_19, type="deviance")
par(mfrow=c(2,2))
plot(model_19)
```



## b. All seasons together
By fitting the model for all the seasons together (which captures an overall positive effect of playing home even though the effect is not present during covid), we obtain the diagnostics displayed in *Figure 3*. They globally have the same shape as the previous ones. Thus, we identify some influential outliers which indicate that the model is not perfect.

```{r, fig.align="center", fig.cap="Figure 3 : Residuals diagnostics for the model considering all seasons"}
par(mfrow=c(2,2))
model_wins <- glm(df_wins_all$number ~ df_wins_all$team_status, family=poisson(link="log"))
plot(model_wins)
```

# 5. Stability 
## a. Promoted and relegated teams
The model we built so far didn't take in account the fact that some teams are relegated and some other promoted from season to season. We discard those teams and refit the Poisson model of part 5.b. The estimated coefficients are as follows 
```{r promoted, echo=FALSE, include = FALSE}
#relegated/promoted effect
A = setdiff(df18$HomeTeam,df19$HomeTeam) #relegated (those in for 18 but not in 19)
B = setdiff(df19$HomeTeam,df18$HomeTeam) #promoted (those in 19 but not in 18)
C = setdiff(df19$HomeTeam, df20$HomeTeam) #relegated
D = setdiff(df20$HomeTeam, df19$HomeTeam) #promoted
E = setdiff(df20$HomeTeam, df21$HomeTeam ) #relegated
f = setdiff(df21$HomeTeam, df20$HomeTeam ) #promoted

names_del <- unique(c(A, B, C, D, E, f)) #we will not take in account the ones that were relegated and promoted

df_wins_before_rel <- df_restr %>% filter(covid_before_after =='Pre-covid')%>%  filter(win == "1")%>% filter(!(team_name%in%names_del))%>%select(team_name, team_status, win)%>% group_by(team_name, team_status)%>%summarise(number=n())%>% mutate(numb_win = number)

df_wins_during_rel <- df_restr %>% filter(covid_before_after =='In-covid')%>%  filter(win == "1")%>% filter(!(team_name%in%names_del))%>%select(team_name, team_status, win)%>% group_by(team_name, team_status)%>%summarise(number=n())%>% mutate(numb_win = number)

df_wins_after_rel <- df_restr %>% filter(covid_before_after =='Post-covid')%>%  filter(win == "1")%>% filter(!(team_name%in%names_del))%>%select(team_name, team_status, win)%>% group_by(team_name, team_status)%>%summarise(number=n())%>% mutate(numb_win = number)

model_wins_before_rel <- glm(df_wins_before_rel$number ~ df_wins_before_rel$team_status, family=poisson(link="log"))
summary(model_wins_before_rel)

model_wins_during_rel <- glm(df_wins_during_rel$number ~ df_wins_during_rel$team_status, family=poisson(link="log"))
summary(model_wins_during_rel)

model_wins_after_rel <- glm(df_wins_after_rel$number ~ df_wins_after_rel$team_status, family=poisson(link="log"))
summary(model_wins_after_rel)

```

*Table 3 : Estimated coefficients of the Poisson model without relegated and promoted teams*

|              | Estimate | Std.Error | t value | Pr(>\|t\|) |
|--------------|----------|-----------|---------|------------|
| Before covid | 0.28     | 0.1       | 2.86    | 0.004      |
| During covid | -0.07    | 0.13      | -0.51   | 0.61       |
| After covid  | 0.29     | 0.13      | 2.29    | 0.02       |



## b. Draw effect
Another limit of our model in 5.b was that we only counted the number of wins, but we could also include the draws as a victory for both teams. By doing this, the coefficients estimation gives


```{r draw effect, echo=FALSE, include = FALSE}
#if we consider that draw is almost like winning 
df_wins_before_bis <- as.data.frame(df_restr %>% filter(covid_before_after =='Pre-covid')%>%  filter(win != "0")%>% select(team_name, team_status, win)%>% group_by(team_name, team_status)%>%summarise(number=n())%>% mutate(numb_win = number))

df_wins_during_bis <- df_restr %>% filter(covid_before_after =='In-covid')%>%  filter(win != "0")%>% select(team_name, team_status, win)%>% group_by(team_name, team_status)%>%summarise(number=n())%>% mutate(numb_win = number)

df_wins_after_bis <- df_restr %>% filter(covid_before_after =='Post-covid')%>%  filter(win != "0")%>% select(team_name, team_status, win)%>% group_by(team_name, team_status)%>%summarise(number=n())%>% mutate(numb_win = number)


model_wins_before_bis <- glm(df_wins_before_bis$number ~ df_wins_before_bis$team_status, family=poisson(link="log"))
summary(model_wins_before_bis)

model_wins_during_bis <- glm(df_wins_during_bis$number ~ df_wins_during_bis$team_status, family=poisson(link="log"))
summary(model_wins_during_bis)

model_wins_after_bis <- glm(df_wins_after_bis$number ~ df_wins_after_bis$team_status, family=poisson(link="log"))
summary(model_wins_after_bis)
```


*Table 4 : Estimated coefficients of the Poisson model taking in account the draws*

|              | Estimate | Std.Error | t value | Pr(>\|t\|) |
|--------------|----------|-----------|---------|------------|
| Before covid | 0.21     | 0.07      | 2.9     | 0.004      |
| During covid | -0.04    | 0.09      | -0.42   | 0.68       |
| After covid  | 0.15     | 0.09      | 1.57    | 0.12       |



# 6. Interpretation
The first model in section 3.a is a first approach to quantify the home advantage taking in account the difference of number of goals. It appears very clearly that there was a home advantage before covid-19 occurred since the coefficient associated to playing home is positive and statistically significant at every usual level. It is also practically significant since its value is 0.27 (95CI = [0.14, 0.4]), which means that a team marks 0.27 more goals when they play at home. During covid-19, the coefficient associated to playing home is not statistically significant and 0 belongs to the confidence interval, which means that there is no effect of playing home rather than playing away. After the closed door period, the effect of playing home becomes again significant statistically (at the 5% level) and practically. 

The second approach to this problem is to consider the number of victories (strictly victory) and model it by a poisson or log-linear regression. The estimated coefficient before covid-19 is again statistically and practically (0.32) significant at the usual levels (5% and 1%). Again, the coefficient associated to the period during covid-19 is not significant so there was no home advantage. Finally, the home advantage reappears after the opening of stadiums the same way we noticed it with the linear regression.



# 7. Discussion
The models built seem plausible and capture the home advantage decrease during covid-19 due to the closed to public stadiums. The diagnostics and stability analysis showed that the model was strong which gives confidence. 

The interesting conclusion of our work is that after covid-19, the home advantage was rebuilt and did not disappear. It is generally of the same magnitude but a bit less important. To explain that decrease, some plausible hypothesis could be the fact that the teams accustomed themselves to empty stadiums, or that the supporters did not act the same way before and after the covid period.

We did not explore the yellow and red cards that were studied by Garicano and Palacios-Huertas [1], which could be also very interesting. We also analyzed in the linear regression model the number of goals but we could have instead analyzed the number of targeted goals. 


# 8. References
[1] Garicano, L. and Palacios-Huerta, I. (2001). "An empirical examination of multilateral bargaining." American Economic Review, 91(4): 828-848.

[2] Nevill, A. M., Newell, S. M., & Gale, S. (1996). Factors associated with home advantage in English and Scottish soccer matches. Journal of Sports Sciences, 14(2), 181-186.

[3] Pollard, R., & Pollard, G. (2005). Home advantage in soccer: Variations in its magnitude and a literature review of the inter-related factors associated with its existence. Journal of Sport Behavior, 28(3), 169-189.